---
title: "Risk_Index_Violence"
---

```{r}
#Hypothesis I: Contestation associated with Violence

library(plm)
library(lmtest)

# Panel data (state-year panel)
df_panel <- pdata.frame(df, index = c("STATE", "YEAR"))


#Panel Data Analysis with Fixed Effects for Year and State
#Total Homicide Rate
fe_total_time <- plm(TOTAL_RATE ~ Contestation_Level + factor(YEAR) + factor(STATE), data = df_panel, model = "within")
summary(fe_total_time)

coeftest(fe_total_time, vcov = vcovHC(fe_total_time, type = "HC1", cluster = "group"))

#Male Homicide Rate
fe_male_time <- plm(MALE_RATE ~ Contestation_Level + factor(YEAR) + factor(STATE), data = df_panel, model = "within")
summary(fe_male_time)

coeftest(fe_male_time, vcov = vcovHC(fe_male_time, type = "HC1", cluster = "group"))

```

```{r}
#Hypothesis II: Risk Index and Homicide Rates as proxy of Cartel Contestation
library(dplyr)
library(tidyverse)
library(sandwich)
library(MASS)

df <- read.csv("EMIF_CBP_corrected_FINAL.csv")
df

df$City <- as.factor(df$City)
df$Female <- as.factor(df$Female)
df$indig <- as.factor(df$indig)
df$eng <- as.factor(df$eng)
df$emp <- as.factor(df$emp)
df$Married <- as.factor(df$Married)
df$Smugg_decision <- as.factor(df$Smugg_decision)
df$rel_border <- as.factor(df$rel_border)
df$Ln_male_hom_rt_corrected <- as.numeric(df$Ln_male_hom_rt_corrected)
df$Ln_male_hom_rt_corrected <- as.numeric(df$Ln_male_hom_rt_corrected)
df$ln_aprehen <- log(df$Aprehen + .001)
# Per 10,000 apprehensions
df$Aprehen_10k <- df$Aprehen / 10000



#Total homicide rate model: Negative Binomial with Fixed Effects for City and Year
m_total <- glm.nb(h_Index ~ Female + indig + Educ + Married + emp + Aprehen_10k +
                         Ln_hom_rt_corrected + Experience + Companions + Age +
                         factor(City) + factor(Year), data = df)

summary(m_total)


#Male Homicide Rate Model: Negative Binomial with Fixed Effects for City and Year
m_male <- glm.nb(h_Index ~ Female + indig + Educ + Married + emp  + Aprehen_10k +
                         Ln_male_hom_rt_corrected + Experience + Companions + Age +
                         factor(City) + factor(Year), data = df)

summary(m_male)

```


```{r}
#Coefficients Plot (Models 1 and 2)
library(broom)
library(dotwhisker)
library(ggplot2)


results <- bind_rows(
  tidy(m_total) %>% mutate(model = "Total Homicide Rate"),
  tidy(m_male)  %>% mutate(model = "Male Homicide Rate")
)


my_terms <- c("Ln_hom_rt_corrected", "Ln_male_hom_rt_corrected",
              "emp1", "Aprehen_10k", "Age", "Female1", "indig1")

results <- results %>% 
  filter(term %in% my_terms) %>%
  mutate(term = recode(term,
                       'Ln_hom_rt_corrected' = 'Total Homicide Rate',
                       'Ln_male_hom_rt_corrected' = 'Male Homicide Rate',
                       'emp1' = 'Employed',
                       'Aprehen_10k' = 'Border Enforcement',
                       'indig1' = 'Indigenous',
                       'Female1' = 'Female'))

# Faceted coefficient plot
dwplot(results, dot_args = list(aes(color = model))) +
  facet_wrap(~model, scales = "free_x") +  # You can add scales="free_x" for balance
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey50", linewidth = 0.4) +
  theme_minimal(base_size = 8) +
  labs(
       x = "Coefficient Estimate (95% CI)",
       y = NULL,
       color = "Model") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 8, face = "bold"),
    panel.spacing = unit(1.2, "cm"),       # <-- Add more space between panels
    axis.text.x = element_text(size = 8), # Ensure labels don’t overlap
    strip.text = element_text(face = "bold", size = 8)
  )

ggsave("effects_migrant_risk_index.png", dpi=600)

```

```{r}
#FE Coefficients Plot

tidy_results <- broom::tidy(m_male) %>%
  filter(term != "(Intercept)")  

predictor_labels <- c(
  Age = "Age",
  Female1 = "Female",
  emp1 = "Employed",
  indig1 = "Indigenous",
  Aprehen_10k = "Border Control",
  Ln_male_hom_rt_corrected = "Male Homicide Rate",
  `factor(City)3` = "Tijuana",
  `factor(City)4` = "Sonoyta",
  `factor(City)10` = "San Luis Río Colorado",
  `factor(City)11` = "Sáric",
  `factor(City)12` = "Sásabe",
  `factor(City)17` = "Ciudad Acuña",
  `factor(City)18` = "Piedras Negras",
  `factor(City)20` = "Matamoros",
  `factor(City)21` = "Nuevo Laredo",
  `factor(City)22` = "Reynosa",
  `factor(City)23` = "Miguel Alemán"
)

variables_to_plot <- c(
  "Age", "emp1", "indig1", "Female1", "Aprehen_10k", "Ln_male_hom_rt_corrected",
  "factor(City)3", "factor(City)4", "factor(City)10", "factor(City)11",
  "factor(City)12", "factor(City)17", "factor(City)18", "factor(City)20",
  "factor(City)21", "factor(City)22", "factor(City)23"
)

plot_data <- tidy_results %>%
  filter(term %in% variables_to_plot) %>% 
  mutate(
    is_city = grepl("City", term),
    is_significant = p.value < 0.05,
    group = case_when(
      is_city & estimate > 0 ~ "Eastern Border Cities",
      is_city & estimate < 0 ~ "Western Border Cities",
      TRUE ~ "Other Variables"
    ),
    label = ifelse(term %in% names(predictor_labels),
                   predictor_labels[term], term)
  )

ggplot(plot_data, aes(x = estimate, y = reorder(label, estimate),
                      xmin = estimate - 1.96 * std.error,
                      xmax = estimate + 1.96 * std.error,
                      color = group)) +
  geom_point(size = 2) +
  geom_errorbarh(height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_color_manual(values = c(
    "Eastern Border Cities" = "firebrick",
    "Western Border Cities" = "steelblue",
    "Other Variables" = "gray70"
  )) +
  labs(
    x = "Coefficient Estimate (95% CI)",
    y = "",
    color = "Group"
  ) +
  theme_minimal(base_size = 6) +
  theme(
    plot.title = element_text(size = 6, face = "bold", hjust = 0.3),
    axis.title = element_text(size = 6),
    axis.text = element_text(size = 6),
    legend.position = "bottom"
  )

ggsave(
  "east_vs_west_fixed_effects.png",
  width = 6, height = 4, units = "in",
  dpi = 600
)

```

```{r}

#Hypothesis III: Regions of Risk
library(ggdendro)

df <- read.csv("EMIF_CBP.csv")

city_df <- df %>%
  dplyr::group_by(City) %>%
  dplyr::summarise(across(dplyr::starts_with("h_"), \(x) mean(x, na.rm = TRUE)))

# Robust scaling
robust_scale <- function(x) (x - median(x, na.rm = TRUE)) / IQR(x, na.rm = TRUE)
city_scaled <- city_df %>%
  dplyr::mutate(across(dplyr::starts_with("h_"), robust_scale))

risk_vars <- dplyr::select(city_scaled, dplyr::starts_with("h_"))
rownames(risk_vars) <- city_df$City

# Distance + hierarchical clustering
dist_matrix <- dist(risk_vars, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")

# Dendrogram data and clusters
dend_data <- ggdendro::dendro_data(hc, type = "rectangle")

k <- 3
clusters <- cutree(hc, k = k)
cluster_color_map <- c("2" = "#FDD835",  
                       "1" = "#FB8C00",  
                       "3" = "#B71C1C") 

p <- ggplot() +
  geom_segment(
    data = dend_data$segments,
    aes(x = x, y = y, xend = xend, yend = yend),
    color = "grey35", linewidth = 0.5
  ) +
  geom_text(
    data = dend_data$labels,
    aes(x = x, y = y - 0.1, label = label,
        color = factor(clusters[hc$order])),
    size = 2.8, vjust = 1
  ) +
  scale_color_manual(values = cluster_color_map, guide = "none") +
  labs(x = NULL, y = "Height") +
  theme_minimal(base_size = 10, base_family = "Times New Roman") +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(size = 10),
    legend.position = "none",
    panel.grid = element_blank(),
    plot.margin = margin(5, 5, 5, 5)
  )

ggsave("Hierarchical_Dendrogram_clean.png", plot = p,
       width = 8, height = 5, dpi = 600)

```

```{r}
library(cluster)
library(factoextra)
library(patchwork)

risk_vars <- city_scaled %>% dplyr::select(starts_with("h_"))

dist_matrix <- dist(risk_vars, method = "euclidean")

# Elbow Method
wss <- map_dbl(1:10, function(k) {
  kmeans(risk_vars, centers = k, nstart = 25)$tot.withinss
})

elbow_df <- tibble(k = 1:10, wss = wss)

# visually chosen elbow (≈3)
elbow_k <- 3  

elbow_plot <- ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line(linewidth = 1, color = "#0072B2") +
  geom_point(size = 2.5, color = "#D55E00") +
  geom_vline(xintercept = elbow_k, linetype = "dashed",
             color = "gray40", linewidth = 0.6) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Number of Clusters", y = "Total Within-Cluster Sum of Squares") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(face = "plain"),
    axis.text = element_text(color = "black"),
    plot.title = element_blank()
  )

# Silhouette Analysis
sil_df <- map_dfr(2:10, function(k) {
  km <- kmeans(risk_vars, centers = k, nstart = 25)
  ss <- silhouette(km$cluster, dist_matrix)
  tibble(k = k, silhouette = mean(ss[, 3]))
})

# find optimal silhouette automatically
opt_k <- sil_df$k[which.max(sil_df$silhouette)]

sil_plot <- ggplot(sil_df, aes(x = k, y = silhouette)) +
  geom_line(linewidth = 1, color = "#009E73") +
  geom_point(size = 2.5, color = "#E69F00") +
  geom_vline(xintercept = opt_k, linetype = "dashed",
             color = "gray40", linewidth = 0.6) +
  scale_x_continuous(breaks = 2:10) +
  labs(x = "Number of Clusters", y = "Average Silhouette Width") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(face = "plain"),
    axis.text = element_text(color = "black"),
    plot.title = element_blank()
  )

combined_plot <- elbow_plot + sil_plot + plot_layout(ncol = 2)

ggsave("Elbow_Silhouette_Combined_dashed.png",
       plot = combined_plot,
       dpi = 600,
       width = 10, height = 4.5, units = "in")

```

```{r}
library(sf)
library(ggplot2)
library(dplyr)
library(readr)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggrepel)

us_states <- rnaturalearth::ne_states(country = "United States of America", returnclass = "sf") %>%
  filter(name %in% c("California", "Arizona", "New Mexico", "Texas"))

cities <- read_csv("border_cities_coords.csv", show_col_types = FALSE)
cluster_names <- c("Risk Level 1", "Risk Level 2", "Risk Level 3")
cities$Cluster <- factor(cities$Cluster, levels = 1:3, labels = cluster_names)

cities_to_label <- cities %>% 
  filter(Border %in% c("Tijuana", "Ciudad Juarez"))

state_labels <- data.frame(
  state = c("Baja California", "Sonora", "Chihuahua", "Coahuila", "Nuevo León", "Tamaulipas"),
  lon   = c(-115, -111, -106.5, -102.5, -100, -98),
  lat   = c(30.7, 29.7, 29, 27.5, 25.5, 25)  # ⬅️ lowered Nuevo León from 26 to 25.3
)

us_labels <- data.frame(
  state = c("CA", "AZ", "NM", "TX"),
  lon   = c(-116.5, -112, -106.3, -99.5),
  lat   = c(33.5, 33, 33, 32)
)

north_states <- c("Baja California", "Sonora", "Chihuahua", "Coahuila", "Nuevo León", "Tamaulipas")
mx_states_north <- mx_states %>% filter(NAME_1 %in% north_states)




# ---- Plot ----
p <- ggplot() +
  borders("world", regions = c("Mexico", "USA"), fill = "gray95", colour = "gray80") +
  
  geom_sf(data = us_states, fill = NA, color = "gray60", linewidth = 0.4) +
  
  geom_sf(data = mx_states_north, fill = NA, color = "gray40", size = 0.5) +
  geom_point(data = cities, 
             aes(x = Longitude, 
                 y = Latitude, fill = Cluster),
             shape = 21,
             size = 2.5,
             color = "black",
             stroke = 0.2) +
  geom_text_repel(
    data = cities_to_label,
    aes(x = Longitude, y = Latitude, label = Border),
    size = 1.5,
    family = "Times",
    box.padding = 0.3,
    point.padding = 0.3,
    segment.color = "gray50",
    segment.size = 0.2,
    min.segment.length = 0
  ) +
  geom_text(data = state_labels, aes(x = lon, y = lat, label = state),
            fontface = "bold", size = 2, color = "white", stroke = 0.4) +
  geom_text(data =state_labels, aes(x = lon, y = lat, label = state),
          fontface = "bold", size = 2, color = "gray30", alpha = 0.9) +
  geom_text(
    data = us_labels,
    aes(x = lon, y = lat, label = state),
    fontface = "italic",
    size = 2,
    color = "gray30",
    alpha = 0.9,
    family = "Times"
  ) +
  scale_fill_manual(values = c("dodgerblue3", "orange", "darkred")) +
  labs(fill = "Risk Cluster") +
  
  annotation_scale(
    location = "bl",
    width_hint = 0.2,
    bar_cols = c("gray10", "white"),
    line_width = 0.25,
    text_cex = 0.4,
    unit_category = "metric",
    style = "ticks",
    breaks = 5
  )  +
  annotation_north_arrow(
    location = "bl",
    which_north = "true",
    pad_x = unit(0.45, "in"),
    pad_y = unit(0.35, "in"),
    height = unit(0.45, "in"),
    width = unit(0.45, "in"),
    style = north_arrow_fancy_orienteering(
      fill = c("black", "white"),    
      text_size = 0,                 
      line_width = 0.25
    )
  ) +
  
  coord_sf(xlim = c(-118, -96), ylim = c(24, 34), expand = FALSE) +
  theme_minimal(base_size = 8) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x  = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks   = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank(),
    legend.text = element_text(size = 7),   # Makes "Risk Level 1" etc. larger
    legend.title = element_text(size = 9, face = "bold") # Makes "Risk Cluster" title larger
  )

ggsave("Border_Cities_Risk_withUSBorders.png", plot = p, width = 7.5, height = 4, dpi = 600)

```
```{r}
library(dplyr)
library(ggplot2)
library(ggdendro)
library(dendextend)

df <- read.csv("EMIF_CBP_corrected.csv")

city_df <- df %>%
  group_by(City) %>%
  summarise(across(starts_with("h_"), ~ mean(.x, na.rm = TRUE)))

robust_scale <- function(x) (x - median(x, na.rm = TRUE)) / IQR(x, na.rm = TRUE)
city_scaled <- city_df %>%
  mutate(across(starts_with("h_"), robust_scale))

risk_vars <- dplyr::select(city_scaled, dplyr::starts_with("h_"))
rownames(risk_vars) <- city_df$City


dist_matrix <- dist(risk_vars, method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")

dend <- as.dendrogram(hc)

k <- 3
clusters <- cutree(hc, k = k)


# Add cluster groups to labels in dend_data
label_order <- dend_data$labels$label
cluster_assignment <- clusters[label_order]

segments <- dend_data$segments
labels_df <- dend_data$labels

segments$cluster <- NA
for (i in seq_len(nrow(segments))) {
  match_idx <- which(abs(labels_df$x - segments$xend[i]) < 0.01)
  if (length(match_idx) == 1) {
    segments$cluster[i] <- cluster_assignment[match_idx]
  }
}

segments$cluster <- zoo::na.locf(segments$cluster, fromLast = TRUE)

# ---- Plot ----
p <- ggplot() +
  geom_segment(
    data = segments,
    aes(x = x, y = y, xend = xend, yend = yend, 
        color = factor(cluster)),
    linewidth = 0.7
  ) +
  geom_text(
    data = labels_df,
    aes(x = x, y = y - 0.1, label = label,
        color = factor(cluster_assignment)),
    size = 3, vjust = 1
  ) +
  scale_color_manual(values = c(
    "1" = "#FB8C00",   # orange
    "2" = "blue1",   # yellow
    "3" = "#B71C1C"    # red
  )) +
  guides(color = "none") +
  labs(x = NULL, y = "Height") +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    plot.margin = margin(5, 5, 5, 5)
  )

ggsave("Hierarchical_Dendrogram_colored.png", plot = p,
       width = 8, height = 5, dpi = 600)




